---
title: "ASD Longitudinal RMD"
author: "Stan West"
date: "2025-04-17"
output: html_document
---

This Markdown summarizes information regarding the data that will be used for a longitudinal analysis of vocabulary acquisition in autistic children

```{r, echo=FALSE,warning=FALSE,message=FALSE}
load("data/ASD_longitudinal_all.Rdata")
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(knitr)
library(kableExtra)
diff_scores <- ASD_total %>%
  select(subjectkey,interview_age) %>%
  group_by(subjectkey) %>%
  mutate(entry = seq.int(1,n(),1)) %>%
  filter(!max(entry) < 2) %>%
  mutate(diff = interview_age - lag(interview_age),
         max_diff = max(diff,na.rm = T))

ASD_total <- ASD_total %>%
  filter(nProduced >= 3 & nProduced <= 600)

```


## Demographic Information
Demographics for kids with between 3 & 600 producetive vocabulary

```{r,echo=FALSE}

print(paste0("Number of unique subjectkeys is ",
             length(unique(ASD_total$subjectkey)),
             " and the number of unique CDI entries is ",nrow(ASD_total %>%
               select(subjectkey,interview_age) %>%
               unique())))

ASD_total %>%
  select(interview_age, nProduced) %>%
  summary()


```



## Longitudinal Analysis
# Kids with longitudinal data in 3 month intervals
Of the 497 CDIs, 241 only have an entry for a single age, removing them from longitudinal consideration. That leaves 256(87 unique) with at least 2 CDIs. There are only three kids with CDIs at the specificity of 1 - 4 months between CDIs.

```{r, echo = FALSE, message=FALSE,warning=FALSE}
diff_scores <- ASD_total %>%
  select(subjectkey,interview_age) %>%
  unique() %>%
  arrange(subjectkey, interview_age) %>%
  group_by(subjectkey) %>%
  mutate(entry = seq.int(1,n(),1)) %>%
  filter(max(entry) >= 2) %>%
  mutate(diff = interview_age - lag(interview_age),
         max_diff = max(diff,na.rm = T),
         n_entry = length(entry)) %>%
  left_join(unique(select(ASD_total,subjectkey,interview_age,nProduced)), by = c("subjectkey","interview_age"))

ggplot(diff_scores, aes(x = diff))+
  geom_histogram()+
  labs(caption = "Only 3 participants with between 1 and 4 months distance between CDIs")+
  geom_vline(xintercept = c(1,4), color = "red",linetype = 2)

ggplot(diff_scores %>%
         filter(subjectkey %in% filter(diff_scores,diff <=4 & diff >= 1)$subjectkey),
       aes(x = interview_age, y = nProduced))+
  geom_point(aes(color = subjectkey))+
  geom_line(aes(color = subjectkey))

```

# Kids with longitudinal data in 12 month intervals
There are 75 kids with CDIs at the specificity of 9 - 15 months between CDIs. 45 of those only have one instance of a gap between 9 and 15 months,
23 have two instances of 9 - 15 month gaps, and 7 have three instances of this.

```{r, echo = FALSE, message=FALSE,warning=FALSE}

df_9_15 <- diff_scores %>%
  filter(subjectkey %in% filter(diff_scores,diff >= 9 & diff <= 15)$subjectkey) %>%
  group_by(subjectkey) %>%
  mutate(is.between = ifelse(diff >= 9 & diff <= 15,TRUE,FALSE),
         n_9_15 = sum(is.between,na.rm = T))

## Kid count
length(unique(df_9_15$subjectkey))

## Count number of 9 - 15 CDI gaps
df_9_15 %>%
  select(subjectkey,n_9_15) %>%
  unique() %>%
  group_by(n_9_15) %>%
  count(n_9_15)


  

df_9_15 %>%
  group_by(n_entry) %>%
  table()

ggplot(diff_scores, aes(x = diff))+
  geom_histogram()+
  labs(caption = "75 with between 9 and 15 months distance between CDIs")+
  geom_vline(xintercept = c(9,15), color = "red",linetype = 2)

ggplot(df_9_12,
     aes(x = interview_age, y = nProduced))+
  geom_point(aes(color = subjectkey))+
  geom_line(aes(color = subjectkey))+
  theme(legend.position = "none")

ggplot(df_9_12 %>%
         mutate(interview_age_cut = cut(interview_age, seq(12, 132,11))),
     aes(x = interview_age, y = nProduced))+
  geom_point(aes(color = subjectkey))+
  geom_line(aes(group = subjectkey, color = is.between))+
  theme(legend.position = "none")+
  labs(caption = "Red connections are 9-15 month gaps between CDIs")

```




