---
title: "ASD Longitudinal RMD"
author: "Stan West"
date: "2025-04-17"
output: pdf_document
---

This Markdown summarizes information regarding the data that will be used for a longitudinal analysis of vocabulary acquisition in autistic children. Here we are interested in vocabulary sizes between 3 & 600 words produced. The numbers reported here differ from our first look because of a couple of reasons: 

  1. I recycled old code from a different project to combine the old and new GUIDs. This code had a part where we were only accepting the "best" CDI record (that with the highest number of words produced across multiple records). This was reducing the number of possible longitudinal data points. Good news, we have more kids with longitudinal data 9 - 15 months between CDIs. 
  
  2. I used an "|" instead of "&" when defining the 2 - 4 window for kids with more granular longitudinal data. This made it seem like we had much more kids with fine-tuned longitudinal data. After fixing this, we only have 6 kids with data at that level of specificity.

```{r, echo=FALSE,warning=FALSE,message=FALSE}
load("data/ASD_longitudinal_all.Rdata")
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(ggExtra)
library(forcats)

ASD_total <- ASD_total %>%
  filter(nProduced >= 3 & nProduced <= 600) %>%
  group_by(subjectkey,interview_age) %>%
  slice_max(nProduced)

```


# Demographic Information
Demographics for kids with between 3 & 600 productive vocabulary. The majority of kids have early interviews with relatively low productive vocabularies.

```{r,echo=FALSE, warning=FALSE,message=FALSE,out.width="50%",fig.align='center'}

print(paste0("Number of unique subjectkeys is ",
             length(unique(ASD_total$subjectkey)),
             " and the number of unique CDI entries is ",nrow(ASD_total %>%
               select(subjectkey,interview_age) %>%
               unique())))

ASD_total %>%
  select(interview_age, nProduced) %>%
  summary()

ggplot(ASD_total %>% select(subjectkey,interview_age,nProduced) %>% unique(), aes(x = interview_age, y = nProduced))+
  geom_density_2d_filled()+
  theme(legend.position = "none")+
  labs(x = "Interview Age",
       y = "# of Words Produced",
       title = "Interview Age by Vocabulary Density Map")




```
\newpage


# Longitudinal Analysis

Of the 628 CDIs, 206 only have an entry for a single age, removing them from longitudinal consideration. That leaves 422(111 unique kids) with at least 2 CDIs. 

```{r, echo = FALSE, message=FALSE,warning=FALSE, fig.show='hold'}

num_tbl <- ASD_total %>% 
  select(subjectkey, interview_age) %>%
  unique() %>%
  arrange(subjectkey, interview_age) %>%
  group_by(subjectkey) %>%
  mutate(entry = seq.int(1,n(),1),
         max_entry = max(entry)) %>%
  group_by(max_entry) %>%
  count(max_entry)



num_tbl_pps <- ASD_total %>% 
  select(subjectkey, interview_age) %>%
  unique() %>%
  arrange(subjectkey, interview_age) %>%
  group_by(subjectkey) %>%
  mutate(entry = seq.int(1,n(),1),
         max_entry = max(entry)) %>%
  select(subjectkey, max_entry) %>%
  group_by(max_entry) %>%
  summarize(n = length(unique(subjectkey)))


tbl_all <- left_join(num_tbl,num_tbl_pps, by = "max_entry") %>%
  rename("n_CDI" = "n.x",
         "n_pp" = "n.y")


kable(tbl_all,booktabs = TRUE, col.names = c("# of CDIs","# of Records", "# of Kids"),caption = "CDI Longitudinal Counts ") %>%
  kable_styling(latex_options = "scale_down",
                full_width = FALSE) %>%
  column_spec(1, bold = TRUE, border_right = TRUE, color = "black")


ggplot(tbl_all, aes(x = max_entry, y = n_pp))+
  geom_col()+
  labs(x = "# of Longitudinal CDIs",
       y = "# of Kids")+
  geom_text(aes(label = c(206,102,162,128,30),vjust = -0.5))+
  annotate("text",x = 4, y = 150, label = "Numbers indicate the amount of longitudinal records, \n bars indicate number of kids.")+
  theme_bw()


```

\newpage

# Kids with longitudinal data in 3 month intervals
There are only 6 kids with between 2 and 4 months between CDIs. This is different from our first look at this because we mistakenly used "|" in our filer instead of "&" when defining our time window. There are 9 total records with between 2 and 4 months in between CDIs, with the majority of those (4) being 4 months apart. 

```{r,echo=FALSE,warning=FALSE,message=FALSE,figures-side, fig.show='hold',out.width="50%"}

diff_scores <- ASD_total %>%
  select(subjectkey,interview_age) %>%
  unique() %>%
  arrange(subjectkey, interview_age) %>%
  group_by(subjectkey) %>%
  mutate(entry = seq.int(1,n(),1)) %>%
  filter(max(entry) >= 2) %>%
  mutate(diff = interview_age - lag(interview_age),
         max_diff = max(diff,na.rm = T),
         n_entry = length(entry)) %>%
  left_join(unique(select(ASD_total,subjectkey,interview_age,nProduced)), by = c("subjectkey","interview_age"))

ggplot(diff_scores, aes(x = diff))+
  geom_histogram()+
  labs(caption = "Only 6 kids with between 2 and 4 months distance between CDIs")+
  geom_vline(xintercept = c(2,4), color = "red",linetype = 2)+
  theme_bw()+
  labs(x = "Time Between CDIs",
       y = "Count")+
  scale_x_continuous(breaks = seq.int(1,40,2))

ggplot(diff_scores %>%
         filter(subjectkey %in% filter(diff_scores,diff <=4 & diff >= 2)$subjectkey),
       aes(x = interview_age, y = nProduced))+
  geom_point(aes(color = subjectkey))+
  geom_line(aes(color = subjectkey))+
  theme_bw()+
  labs(x = "Interview Age",
       y = "# of Words Produced")

diff_scores %>%
  filter(subjectkey %in% filter(diff_scores,diff <=4 & diff >= 2)$subjectkey) %>%
  select(subjectkey, interview_age, diff, nProduced) %>%
  kable()

```


\newpage

# Kids with longitudinal data in 12 month intervals
There are 111 kids with CDIs at the specificity of 9 - 15 months between CDIs. Below are tables and figures showing the distribution of time between CDIs both across participants and how many individual participants have multiple CDIs within the 9 - 15 time range. The majority of CDIs that meet the 9 - 15 month criteria fall in the 12-month time between CDIs. Additionally, the majority (79 of 111) only have one instance of records that are 9 - 15 months apart. 

```{r,echo=FALSE,warning=FALSE,message=FALSE}

df_9_15 <- diff_scores %>%
  filter(subjectkey %in% filter(diff_scores,diff >= 9 & diff <= 15)$subjectkey) %>%
  group_by(subjectkey) %>%
  mutate(is.between = diff >= 9 & diff <= 15,
         n_9_15 = sum(is.between,na.rm = T))

## Kid count

print(paste0("There are ",length(unique(df_9_15$subjectkey))," unique kids with CDIs 9 - 15 months between CDIs"))

print(paste0("There are ",nrow(df_9_15 %>% filter(diff >= 9 & diff <= 15))," cases of CDIs being between 9 and 15 months apart. The counts are shown below."))

## Count number of 9 - 15 CDI gaps
tbl_counts <- df_9_15 %>%
  filter(diff >= 9 & diff <= 15) %>%
  select(subjectkey,n_9_15, diff) %>%
  group_by(diff) %>%
  count(diff) 

kable(tbl_counts,col.names = c("Month Latency","Count"),caption = "CDI Latencies Counts")

tbl_n_cont <- df_9_15 %>%
  select(subjectkey,n_9_15) %>%
  unique() %>%
  group_by(n_9_15) %>%
  count(n_9_15)

kable(tbl_n_cont,col.names = c("# of Contributions","Count"),caption = "Multiple Contributions Counts: Note: 1 contribution = 2 CDIs 9-15 months apart") 


  

```

\newpage

# Histograms
## Graphical representation of tables above. Plot on the bottom: red lines reflect the 9 - 15 month cutoffs.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
ggplot(tbl_counts, aes( x = diff, y = n))+
  geom_col()+
  scale_x_continuous(breaks = seq.int(9,15,1))+
  labs(x = "Time Between CDIs",
       y = "Count",
       title = "Histogram of Times Between CDIs (9-15)")+
  theme_bw()

ggplot(diff_scores, aes(x = diff))+
  geom_histogram()+
  labs(caption = "111 with between 9 and 15 months distance between CDIs")+
  geom_vline(xintercept = c(9,15), color = "red",linetype = 2)+
  scale_x_continuous(breaks = seq.int(1,40,2))+
  labs(x = "Time Between CDIs",
       y = "Count",
       title =  "Histogram of Times Between CDIs (All)")+
  theme_bw()


```


## Distributions of first interviews

```{r,message=FALSE,echo=FALSE,warning=FALSE}

ggplot(df_9_15 %>%
         filter(entry == 1),
       aes(x = interview_age))+
  geom_histogram()+
  scale_x_continuous(breaks = seq.int(12,132,11))+
  theme_bw()+
  labs(x = "Interview Age",
       y = "Count",
       title = "Distribution of First Interviews")

ggplot(df_9_15 %>%
         filter(entry == 1) %>%
         mutate(n_entry = factor(n_entry, levels = c(2,3,4,5), labels = c(2,3,4,5))),
       aes(x = interview_age, fill = n_entry))+
  geom_histogram(binwidth = 4)+
  scale_x_continuous(breaks = seq.int(12,132,11))+
  theme_bw()+
  labs(x = "Interview Age",
       y = "Count",
       title = "Distribution of First Interviews")+
  facet_wrap(~n_entry)

```




\newpage

# Spaghetti Plots
Plots showing the progression of productive vocabulary over interview age. The majority of kids show increases in productive vocabulary over time, but this is not always the case. The next page digs deeper into those cases. 


```{r,echo=FALSE,warning=FALSE,message=FALSE,figures-side3,out.height="50%",fig.show='hold'}


ggplot(df_9_15,
     aes(x = interview_age, y = nProduced))+
  geom_point(aes(color = subjectkey))+
  geom_line(aes(color = subjectkey))+
  theme_bw()+
  theme(legend.position = "none")+
  labs(x = "Interview Age",
       y = "# of Words Produced")

ggplot(df_9_15 %>%
         mutate(interview_age_cut = cut(interview_age, seq(12, 132,11))),
     aes(x = interview_age, y = nProduced))+
  geom_point(aes(color = subjectkey))+
  geom_line(aes(group = subjectkey, color = is.between))+
  labs(caption = "Red connections are 9-15 month gaps between CDIs")+
  theme_bw()+
  scale_x_continuous(breaks = seq.int(12,132,11))+
  theme(legend.position = "none")+
  labs(x = "Interview Age",
       y = "# of Words Produced")
```

\newpage

# Inconsistent gains in vocabulary over time
Some of the kids show drops in vocabulary size between CDI administrations. There are 30 kids who have longitudinal CDIs that decrease at some point in development. These may be considerations for removal as they do not follow the pattern of development of the majority of the kids. It is impossible to know if this is due to bad reporting or legitimate losses in productive vocacbulary.


```{r,echo=FALSE, message=FALSE,warning=FALSE}
diff_scores_dec <- diff_scores %>%
  group_by(subjectkey) %>%
  mutate(decrease = ifelse(nProduced < lag(nProduced), TRUE,FALSE)) %>%
  mutate(decrease = ifelse(is.na(decrease),FALSE,decrease)) %>%
  filter(subjectkey %in% .[.$decrease == TRUE,]$subjectkey) %>%
  mutate(dec_group = ifelse(decrease == TRUE,
                            "A",
                            ifelse(lead(decrease) == TRUE,
                                   "A",
                                   NA))) %>%
  mutate(dec_group = ifelse(is.na(dec_group), "B",dec_group),
         dec_group = as.factor(dec_group))


diff_scores_dec <- diff_scores %>%
  group_by(subjectkey) %>%
  mutate(decrease = ifelse(nProduced < lag(nProduced), TRUE,FALSE)) %>%
  mutate(decrease = ifelse(is.na(decrease),FALSE,decrease)) %>%
  filter(subjectkey %in% .[.$decrease == TRUE,]$subjectkey) %>%
  mutate(color = nProduced > lead(nProduced))


ggplot(diff_scores_dec,
     aes(x = interview_age, y = nProduced))+
  geom_path(aes(group = subjectkey, color = color))+
  scale_color_manual(values = c("grey","red"))+
  geom_point(aes(fill = subjectkey))+
  labs(caption = "Red indicates decreases in words produced over CDIs")+
  theme_bw()+
  theme(legend.position = "none")+
  labs(x = "Interview Age",
       y = "# of Words Produced")



diff_scores_dec %>%
  group_by(subjectkey) %>%
  mutate(neg_diffs = lead(nProduced) - nProduced) %>%
  filter(neg_diffs < 0) %>%
  ungroup() %>%
  summarize(m_neg_diffs = mean(neg_diffs),
            sd_neg_diffs = sd(neg_diffs),
            range = range(neg_diffs)) %>%
  kable(col.names = c("Mean Vocabulary Decreases", "SD", "Range"))
```


## Top 40 words not consistently reported across CDIs.

```{r,echo=FALSE, message=FALSE,warning=FALSE}


## Work on computing words that are not consistently reported between CDIs within participants.

first_vocabs <- ASD_total %>%
  filter(subjectkey %in% df_9_15$subjectkey) %>%
  group_by(subjectkey) %>%
  select(subjectkey,interview_age, CDI_Metadata_compatible,Produces, nProduced) %>%
  left_join(select(df_9_15,subjectkey,interview_age,entry), by = c("subjectkey","interview_age")) %>%
  filter(entry == 1, Produces == TRUE) %>%
  mutate(in_1 = ifelse(entry == 1 & Produces == TRUE,1,0))

Itmlvl_not_maintained <- ASD_total %>%
  filter(subjectkey %in% df_9_15$subjectkey) %>%
  group_by(subjectkey) %>%
  select(subjectkey,interview_age, CDI_Metadata_compatible,Produces, nProduced) %>%
  left_join(select(df_9_15,subjectkey,interview_age,entry), by = c("subjectkey","interview_age")) %>%
  group_by(subjectkey,CDI_Metadata_compatible) %>%
  left_join(select(first_vocabs,subjectkey,CDI_Metadata_compatible,in_1))  %>%
  mutate(in_1  = ifelse(is.na(in_1),0,1),
         not_subsequent = ifelse(in_1 == 1 & Produces == FALSE, TRUE, FALSE)) %>%
  filter(!entry == 1,in_1 == 1) %>%
  left_join(ASD_total %>% select(CDI_Metadata_compatible, category) %>% unique())

Itm_lvl_summ <- Itmlvl_not_maintained %>%
  group_by(CDI_Metadata_compatible) %>%
  summarize(sum_itm = sum(not_subsequent))

ggplot(Itm_lvl_summ %>%
         slice_max(sum_itm,n = 40) %>%
         arrange(desc(sum_itm)) %>%
         left_join(ASD_total %>%ungroup() %>% select(CDI_Metadata_compatible,category) %>% unique()), aes(y = reorder(CDI_Metadata_compatible,sum_itm), x = sum_itm, fill = category))+
  geom_bar(stat="identity")+
  theme_bw()+
  labs(y = "Word",
       x = "# of Omissions From First CDI")




```

Here, we are looking at item-level distributions of the top 50 words that were reported as being known in the first CDI administration, but not reported as being known in at least one of the subsequent administrations. The bars are filled by CDI category. Values on the x-axis report the number of times the word was reported on the first administration of the CDI but not later CDIs.


## Overall distribution of omissions across categories

```{r,echo=FALSE, message=FALSE,warning=FALSE}

x <- Itm_lvl_summ %>%
         left_join(ASD_total %>%
                     ungroup() %>% 
                     select(CDI_Metadata_compatible,category) %>% 
                     unique())%>%
  group_by(category) %>%
  summarize(sum = sum(sum_itm)) %>% 
  arrange(sum)


ggplot(Itm_lvl_summ %>%
         left_join(ASD_total %>%
                     ungroup() %>% 
                     select(CDI_Metadata_compatible,category) %>% 
                     unique()) %>%
         mutate(category = factor(category, x$category))
         , aes(y = category, x = sum_itm, fill = category))+
  geom_bar(stat="identity")+
  theme_bw()+
  labs(y = "Category",
       x = "# of Omissions From First CDI")+
  theme(legend.position = "none")





```

The majority of words not being included in subsequent CDIs are action words, followed by animals and food and drinks.




