---
title: "ASD Longitudinal RMD"
author: "Stan West"
date: "2025-04-17"
output: pdf_document
---

This Markdown summarizes information regarding the data that will be used for a longitudinal analysis of vocabulary acquisition in autistic children

```{r, echo=FALSE,warning=FALSE,message=FALSE}
load("data/ASD_longitudinal_all.Rdata")
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(knitr)
library(kableExtra)

ASD_total <- ASD_total %>%
  filter(nProduced >= 3 & nProduced <= 600) %>%
  group_by(subjectkey,interview_age) %>%
  slice_max(nProduced)

```


## Demographic Information
Demographics for kids with between 3 & 600 productive vocabulary

```{r,echo=FALSE, warning=FALSE,message=FALSE}

print(paste0("Number of unique subjectkeys is ",
             length(unique(ASD_total$subjectkey)),
             " and the number of unique CDI entries is ",nrow(ASD_total %>%
               select(subjectkey,interview_age) %>%
               unique())))

ASD_total %>%
  select(interview_age, nProduced) %>%
  summary()


```



## Longitudinal Analysis

Of the 628 CDIs, 206 only have an entry for a single age, removing them from longitudinal consideration. That leaves 422(111 unique kids) with at least 2 CDIs. 

```{r, echo = FALSE, message=FALSE,warning=FALSE}

num_tbl <- ASD_total %>% 
  select(subjectkey, interview_age) %>%
  unique() %>%
  arrange(subjectkey, interview_age) %>%
  group_by(subjectkey) %>%
  mutate(entry = seq.int(1,n(),1),
         max_entry = max(entry)) %>%
  group_by(max_entry) %>%
  count(max_entry)



num_tbl_pps <- ASD_total %>% 
  select(subjectkey, interview_age) %>%
  unique() %>%
  arrange(subjectkey, interview_age) %>%
  group_by(subjectkey) %>%
  mutate(entry = seq.int(1,n(),1),
         max_entry = max(entry)) %>%
  select(subjectkey, max_entry) %>%
  group_by(max_entry) %>%
  summarize(n = length(unique(subjectkey)))


tbl_all <- left_join(num_tbl,num_tbl_pps, by = "max_entry") %>%
  rename("n_CDI" = "n.x",
         "n_pp" = "n.y")


kable(tbl_all,booktabs = TRUE, col.names = c("# of CDIs","# of Records", "# of Kids"),caption = "CDI Longitudinal Counts ") %>%
  kable_styling(latex_options = "scale_down",
                full_width = FALSE) %>%
  column_spec(1, bold = TRUE, border_right = TRUE, color = "black")



```


\newpage

## Kids with longitudinal data in 3 month intervals

```{r,echo=FALSE,warning=FALSE,message=FALSE,figures-side, fig.show='hold',out.width="50%"}

diff_scores <- ASD_total %>%
  select(subjectkey,interview_age) %>%
  unique() %>%
  arrange(subjectkey, interview_age) %>%
  group_by(subjectkey) %>%
  mutate(entry = seq.int(1,n(),1)) %>%
  filter(max(entry) >= 2) %>%
  mutate(diff = interview_age - lag(interview_age),
         max_diff = max(diff,na.rm = T),
         n_entry = length(entry)) %>%
  left_join(unique(select(ASD_total,subjectkey,interview_age,nProduced)), by = c("subjectkey","interview_age"))

ggplot(diff_scores, aes(x = diff))+
  geom_histogram()+
  labs(caption = "Only 6 participants with between 2 and 4 months distance between CDIs")+
  geom_vline(xintercept = c(2,4), color = "red",linetype = 2)+
  theme_bw()+
  labs(x = "Time Between CDIs",
       y = "Count")+
  scale_x_continuous(breaks = seq.int(1,40,2))

ggplot(diff_scores %>%
         filter(subjectkey %in% filter(diff_scores,diff <=4 & diff >= 2)$subjectkey),
       aes(x = interview_age, y = nProduced))+
  geom_point(aes(color = subjectkey))+
  geom_line(aes(color = subjectkey))+
  theme_bw()+
  labs(x = "Interview Age",
       y = "# of Words Produced")

diff_scores %>%
  filter(subjectkey %in% filter(diff_scores,diff <=4 & diff >= 2)$subjectkey) %>%
  select(subjectkey, interview_age, diff, nProduced) %>%
  kable()

```


\newpage

# Kids with longitudinal data in 12 month intervals
There are 111 kids with CDIs at the specificity of 9 - 15 months between CDIs. Below are tables and figures showing the distribution of time between CDIs both across participants and how many individual participants have multiple CDIs within the 9 - 15 time range. 

```{r,echo=FALSE,warning=FALSE,message=FALSE}

df_9_15 <- diff_scores %>%
  filter(subjectkey %in% filter(diff_scores,diff >= 9 & diff <= 15)$subjectkey) %>%
  group_by(subjectkey) %>%
  mutate(is.between = ifelse(diff >= 9 & diff <= 15,TRUE,FALSE),
         n_9_15 = sum(is.between,na.rm = T))

## Kid count

print(paste0("There are ",length(unique(df_9_15$subjectkey))," unique kids with CDIs 9 - 15 months between CDIs"))

print(paste0("There are ",nrow(df_9_15 %>% filter(diff >= 9 & diff <= 15))," cases of CDIs being between 9 and 15 months apart. The counts are shown below."))

## Count number of 9 - 15 CDI gaps
tbl_counts <- df_9_15 %>%
  filter(diff >= 9 & diff <= 15) %>%
  select(subjectkey,n_9_15, diff) %>%
  group_by(diff) %>%
  count(diff) 

kable(tbl_counts,col.names = c("Month Latency","Count"),caption = "CDI Latencies Counts") 

tbl_n_cont <- df_9_15 %>%
  select(subjectkey,n_9_15) %>%
  unique() %>%
  group_by(n_9_15) %>%
  count(n_9_15)

kable(tbl_n_cont,col.names = c("# of Contributions","Count"),caption = "Multiple Contributions Counts") 


  

```

## Histograms

```{r,echo=FALSE,warning=FALSE,message=FALSE,figures-side2, fig.show='hold',out.width="50%",fig.cap="Histograms of 9 - 15 months between CDIs"}
ggplot(tbl_counts, aes( x = diff, y = n))+
  geom_col()+
  scale_x_continuous(breaks = seq.int(9,15,1))+
  labs(x = "Time Between CDIs",
       y = "Count",
       title = "Histogram of Times Between CDIs (9-15)")+
  theme_bw()

ggplot(diff_scores, aes(x = diff))+
  geom_histogram()+
  labs(caption = "111 with between 9 and 15 months distance between CDIs")+
  geom_vline(xintercept = c(9,15), color = "red",linetype = 2)+
  scale_x_continuous(breaks = seq.int(1,40,2))+
  labs(x = "Time Between CDIs",
       y = "Count",
       title =  "Histogram of Times Between CDIs (All)")+
  theme_bw()

```

\newpage

## Spaghetti Plots

```{r,echo=FALSE,warning=FALSE,message=FALSE,figures-side3, fig.show='hold',out.width="50%",fig.cap="Spaghetti plots of longitudinal vocabulary growth"}


ggplot(df_9_15,
     aes(x = interview_age, y = nProduced))+
  geom_point(aes(color = subjectkey))+
  geom_line(aes(color = subjectkey))+
  theme_bw()+
  theme(legend.position = "none")+
  labs(x = "Interview Age",
       y = "# of Words Produced")

ggplot(df_9_15 %>%
         mutate(interview_age_cut = cut(interview_age, seq(12, 132,11))),
     aes(x = interview_age, y = nProduced))+
  geom_point(aes(color = subjectkey))+
  geom_line(aes(group = subjectkey, color = is.between))+
  labs(caption = "Red connections are 9-15 month gaps between CDIs")+
  theme_bw()+
  theme(legend.position = "none")+
  labs(x = "Interview Age",
       y = "# of Words Produced")
```


