---
title: "ASD Longitudinal RMD"
author: "Stan West"
date: "2025-04-17"
output: pdf_document
---

This Markdown summarizes information regarding the data that will be used for a longitudinal analysis of vocabulary acquisition in autistic children. Here we are interested in vocabulary sizes between 3 & 600 words produced. The numbers reported here differ from our first look because of a couple of reasons: 

  1. I recycled old code from a different project to combine the old and new GUIDs. This code had a part where we were only accepting the "best" CDI record (that with the highest number of words produced across multiple records). This was reducing the number of possible longitudinal data points. Good news, we have more kids with longitudinal data 9 - 15 months between CDIs. 
  
  2. I used an "|" instead of "&" when defining the 2 - 4 window for kids with more granular longitudinal data. This made it seem like we had much more kids with fine-tuned longitudinal data. After fixing this, we only have 6 kids with data at that level of specificity.

```{r, echo=FALSE,warning=FALSE,message=FALSE}
load("data/ASD_longitudinal_all.Rdata")
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(ggExtra)

ASD_total <- ASD_total %>%
  filter(nProduced >= 3 & nProduced <= 600) %>%
  group_by(subjectkey,interview_age) %>%
  slice_max(nProduced)

```


## Demographic Information
Demographics for kids with between 3 & 600 productive vocabulary

```{r,echo=FALSE, warning=FALSE,message=FALSE,out.width="50%",fig.align='center'}

print(paste0("Number of unique subjectkeys is ",
             length(unique(ASD_total$subjectkey)),
             " and the number of unique CDI entries is ",nrow(ASD_total %>%
               select(subjectkey,interview_age) %>%
               unique())))

ASD_total %>%
  select(interview_age, nProduced) %>%
  summary()

ggplot(ASD_total %>% select(subjectkey,interview_age,nProduced) %>% unique(), aes(x = interview_age, y = nProduced))+
  geom_density_2d_filled()+
  theme(legend.position = "none")+
  labs(x = "Interview Age",
       y = "# of Words Produced",
       title = "Interview Age by Vocabulary Density Map")




```
\newpage


## Longitudinal Analysis

Of the 628 CDIs, 206 only have an entry for a single age, removing them from longitudinal consideration. That leaves 422(111 unique kids) with at least 2 CDIs. 

```{r, echo = FALSE, message=FALSE,warning=FALSE, fig.show='hold'}

num_tbl <- ASD_total %>% 
  select(subjectkey, interview_age) %>%
  unique() %>%
  arrange(subjectkey, interview_age) %>%
  group_by(subjectkey) %>%
  mutate(entry = seq.int(1,n(),1),
         max_entry = max(entry)) %>%
  group_by(max_entry) %>%
  count(max_entry)



num_tbl_pps <- ASD_total %>% 
  select(subjectkey, interview_age) %>%
  unique() %>%
  arrange(subjectkey, interview_age) %>%
  group_by(subjectkey) %>%
  mutate(entry = seq.int(1,n(),1),
         max_entry = max(entry)) %>%
  select(subjectkey, max_entry) %>%
  group_by(max_entry) %>%
  summarize(n = length(unique(subjectkey)))


tbl_all <- left_join(num_tbl,num_tbl_pps, by = "max_entry") %>%
  rename("n_CDI" = "n.x",
         "n_pp" = "n.y")


kable(tbl_all,booktabs = TRUE, col.names = c("# of CDIs","# of Records", "# of Kids"),caption = "CDI Longitudinal Counts ") %>%
  kable_styling(latex_options = "scale_down",
                full_width = FALSE) %>%
  column_spec(1, bold = TRUE, border_right = TRUE, color = "black")


ggplot(tbl_all, aes(x = max_entry, y = n_pp))+
  geom_col()+
  labs(x = "# of Longitudinal CDIs",
       y = "# of Kids")+
  geom_text(aes(label = c(206,102,162,128,30),vjust = -0.5))+
  annotate("text",x = 4, y = 150, label = "Numbers indicate the amount of longitudinal records, \n bars indicate number of kids.")+
  theme_bw()


```


\newpage

## Kids with longitudinal data in 3 month intervals
There are only 6 kids with between 2 and 4 months between CDIs. This is different from our first look at this because we mistakenly used "|" in our filer instead of "&" when defining our time window.

```{r,echo=FALSE,warning=FALSE,message=FALSE,figures-side, fig.show='hold',out.width="50%"}

diff_scores <- ASD_total %>%
  select(subjectkey,interview_age) %>%
  unique() %>%
  arrange(subjectkey, interview_age) %>%
  group_by(subjectkey) %>%
  mutate(entry = seq.int(1,n(),1)) %>%
  filter(max(entry) >= 2) %>%
  mutate(diff = interview_age - lag(interview_age),
         max_diff = max(diff,na.rm = T),
         n_entry = length(entry)) %>%
  left_join(unique(select(ASD_total,subjectkey,interview_age,nProduced)), by = c("subjectkey","interview_age"))

ggplot(diff_scores, aes(x = diff))+
  geom_histogram()+
  labs(caption = "Only 6 kids with between 2 and 4 months distance between CDIs")+
  geom_vline(xintercept = c(2,4), color = "red",linetype = 2)+
  theme_bw()+
  labs(x = "Time Between CDIs",
       y = "Count")+
  scale_x_continuous(breaks = seq.int(1,40,2))

ggplot(diff_scores %>%
         filter(subjectkey %in% filter(diff_scores,diff <=4 & diff >= 2)$subjectkey),
       aes(x = interview_age, y = nProduced))+
  geom_point(aes(color = subjectkey))+
  geom_line(aes(color = subjectkey))+
  theme_bw()+
  labs(x = "Interview Age",
       y = "# of Words Produced")

diff_scores %>%
  filter(subjectkey %in% filter(diff_scores,diff <=4 & diff >= 2)$subjectkey) %>%
  select(subjectkey, interview_age, diff, nProduced) %>%
  kable()

```


\newpage

# Kids with longitudinal data in 12 month intervals
There are 111 kids with CDIs at the specificity of 9 - 15 months between CDIs. Below are tables and figures showing the distribution of time between CDIs both across participants and how many individual participants have multiple CDIs within the 9 - 15 time range. The majority of CDIs that meet the 9 - 15 month criteria fall in the 12-month time between CDIs. Additionally, the majority (79 of 111) only have one instance of records that are 9 - 15 months apart. 

```{r,echo=FALSE,warning=FALSE,message=FALSE}

df_9_15 <- diff_scores %>%
  filter(subjectkey %in% filter(diff_scores,diff >= 9 & diff <= 15)$subjectkey) %>%
  group_by(subjectkey) %>%
  mutate(is.between = ifelse(diff >= 9 & diff <= 15,TRUE,FALSE),
         n_9_15 = sum(is.between,na.rm = T))

## Kid count

print(paste0("There are ",length(unique(df_9_15$subjectkey))," unique kids with CDIs 9 - 15 months between CDIs"))

print(paste0("There are ",nrow(df_9_15 %>% filter(diff >= 9 & diff <= 15))," cases of CDIs being between 9 and 15 months apart. The counts are shown below."))

## Count number of 9 - 15 CDI gaps
tbl_counts <- df_9_15 %>%
  filter(diff >= 9 & diff <= 15) %>%
  select(subjectkey,n_9_15, diff) %>%
  group_by(diff) %>%
  count(diff) 

kable(tbl_counts,col.names = c("Month Latency","Count"),caption = "CDI Latencies Counts")

tbl_n_cont <- df_9_15 %>%
  select(subjectkey,n_9_15) %>%
  unique() %>%
  group_by(n_9_15) %>%
  count(n_9_15)

kable(tbl_n_cont,col.names = c("# of Contributions","Count"),caption = "Multiple Contributions Counts: Note: 1 contribution = 2 CDIs 9-15 months apart") 


  

```



## Histograms
Graphical representation of tables above. Plot on the right: red lines reflect the 9 - 15 month cutoffs.

```{r,echo=FALSE,warning=FALSE,message=FALSE,figures-side2, fig.show='hold',out.width="50%",fig.cap="Histograms of 9 - 15 months between CDIs"}
ggplot(tbl_counts, aes( x = diff, y = n))+
  geom_col()+
  scale_x_continuous(breaks = seq.int(9,15,1))+
  labs(x = "Time Between CDIs",
       y = "Count",
       title = "Histogram of Times Between CDIs (9-15)")+
  theme_bw()

ggplot(diff_scores, aes(x = diff))+
  geom_histogram()+
  labs(caption = "111 with between 9 and 15 months distance between CDIs")+
  geom_vline(xintercept = c(9,15), color = "red",linetype = 2)+
  scale_x_continuous(breaks = seq.int(1,40,2))+
  labs(x = "Time Between CDIs",
       y = "Count",
       title =  "Histogram of Times Between CDIs (All)")+
  theme_bw()

```

\newpage

## Spaghetti Plots

```{r,echo=FALSE,warning=FALSE,message=FALSE,figures-side3, fig.show='hold',out.width="50%",fig.cap="Spaghetti plots of longitudinal vocabulary growth"}


ggplot(df_9_15,
     aes(x = interview_age, y = nProduced))+
  geom_point(aes(color = subjectkey))+
  geom_line(aes(color = subjectkey))+
  theme_bw()+
  theme(legend.position = "none")+
  labs(x = "Interview Age",
       y = "# of Words Produced")

ggplot(df_9_15 %>%
         mutate(interview_age_cut = cut(interview_age, seq(12, 132,11))),
     aes(x = interview_age, y = nProduced))+
  geom_point(aes(color = subjectkey))+
  geom_line(aes(group = subjectkey, color = is.between))+
  labs(caption = "Red connections are 9-15 month gaps between CDIs")+
  theme_bw()+
  theme(legend.position = "none")+
  labs(x = "Interview Age",
       y = "# of Words Produced")
```


