---
title: "Minimally Speaking Breakdown"
author: "Stan West"
date: "2025-07-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(readxl)
library(ggplot2)

d <- read_xlsx("data/MinmallySpeaking_SHW.xlsx") |>
  select(subjectkey,interview_age, nProduced, form) |>
  na.omit()

```

Data were collected from NDAR and classified as minimally speaking using SET CRITERIA. Here, we split the data according to age of CDI administration and are attempting to match 2-4, 4-6, and 6-8 year age bins. Networks based on these data may reveal different learning mechanisms in minimally speaking children as observed through network structure.


# Everyone

```{r}
ggplot(d, aes(x = nProduced))+
  geom_histogram(stat = "density")+
  facet_grid(~form)
```



## look at older kids  6-8 first then move to the 4-6 then 2 - 4. What form and what are the vocabulary distributions
# Bin 1: 72 - 96 months
```{r}
oldest_group <- d |>
  filter(interview_age >= 72 & interview_age < 96) |>
  group_by(subjectkey) |>
  arrange(subjectkey,interview_age) |>
  mutate(months_between = interview_age - lag(interview_age))


oldest_group |>
  group_by(subjectkey) |>
  filter(length(interview_age) > 1)


ggplot(oldest_group, aes(x = nProduced))+
  geom_histogram(stat = "density")+
  facet_grid(~form)
```

