---
title: "MinimallySpeaking-Longitudinal"
author: "Stan West"
date: "2025-07-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(dplyr)
library(readxl)
library(ggplot2)
library(WeightIt)
library(cobalt)
library(MatchIt)
library(kableExtra)
library(jtools)
library(TriMatch)
library(tidyverse)
library(purrr)


keep_minmax <- function(x){
  r <- x |>
    group_by(subjectkey) |>
    reframe(interview_age = range(interview_age))
  x <- semi_join(x,r, by = c("subjectkey", "interview_age"))
  return(x)
}



d <- read_xlsx("data/MinmallySpeaking_SHW.xlsx") |>
  select(subjectkey,interview_age, nProduced, form) |>
  na.omit() |>
  filter(interview_age <=126 & interview_age > 72) |>
  mutate(group = "old") |>
  arrange(subjectkey,interview_age) |>
  group_by(subjectkey) |>
  arrange(subjectkey,interview_age) |>
  mutate(CDI_dist = interview_age-lag(interview_age)) |>
  filter(CDI_dist >= 9 & CDI_dist <=15 | is.na(CDI_dist)) |>
  mutate(year_dist = abs(CDI_dist - 12)) |>
  filter(!c(subjectkey == "NDARFZ559LLV" & interview_age == 107),
         !c(subjectkey == "NDARWB281CXV" & interview_age == 110)) |>
  mutate(n_entry = length(interview_age))




load("data/ASD_all_GM.Rdata")
ASD_all_filt <- ASD_all |>
  filter(!subjectkey %in% d$subjectkey & interview_age < 72 & interview_age >= 24) |>
  select(subjectkey, interview_age, nProduced, form) |>
  unique() |>
  mutate(group = ifelse(interview_age >= 24 & interview_age < 48, "yng", "mdl"))

ASD_all_filt |>
  select(subjectkey,group) |>
  unique() |>
  group_by(group) |>
  summarize(n = n())

ASD_all_filt <- ASD_all_filt |>
  anti_join(ASD_all |>filter(form == "WG" , interview_age >= 19 , nProduced > 96)) |>
  anti_join(ASD_all |>filter(form == "WS" , interview_age >= 31 , nProduced > 564))

ASD_all_filt |>
  select(subjectkey,group) |>
  unique() |>
  group_by(group) |>
  summarize(n = length(subjectkey))

ASD_all_filt |>
  group_by(subjectkey,group) |>
  mutate(n_entry = length(interview_age)) |>
  select(subjectkey,group,n_entry) |>
  unique() |>
  group_by(group,n_entry) |>
  summarize(n = length(n_entry)) 
  

yng_subs <- ASD_all_filt[ASD_all_filt$group == "yng",] |>
  filter(!subjectkey %in% ASD_all_filt$subjectkey[ASD_all_filt$group == "mdl"]) 

ASD_all_filt <- ASD_all_filt |>
  filter(!group == "yng") |>
  rbind(yng_subs) |>
  group_by(subjectkey) |>
  arrange(subjectkey,interview_age) |>
  mutate(CDI_dist = interview_age - lag(interview_age)) |>
  filter(CDI_dist >= 9 & CDI_dist <=15 | is.na(CDI_dist)) |>
  mutate(n_entry = length(interview_age))

ASD_all_filt |>
  select(subjectkey,group) |>
  unique() |>
  group_by(group) |>
  summarize(n = length(subjectkey))

all_subs <- rbind(d,ASD_all_filt) |>
  mutate(group = factor(group, levels = c("yng","mdl","old")))

all_subs |>
  select(subjectkey,n_entry, group,CDI_dist) |>
  unique() |>
  group_by(group,n_entry) |>
  summarize(count = length(n_entry),
            mean_CDI_dist = mean(CDI_dist,na.rm = T))
  
```

Data were collected from NDAR and classified as minimally speaking using SET CRITERIA. Here, we split the data according to age of CDI administration and are attempting to match 2-4, 4-6, and 6-8 year age bins. Networks based on these data may reveal different learning mechanisms in minimally speaking children as observed through network structure. 

Data used here has first been filtered in several ways: 

1. The oldest group (also the minimally speaking group) was filtered to include children from 72 months to 126 months. If a child has more than one CDI administration, we kept the two administrations that were closest to a year apart (within a 9 - 15 month range; $n_{oldest}$ = 51).
2. The middle and youngest groups only include children from 24 - 47 and 48 - 71 months, respectively and does not include any children also in the oldest group ($n_{youngest}$ = 149,$n_{middle}$ = 55).
3. The middle and youngest groups were then filtered to include appropriate entries given the standards set by the CDI (exclude WG forms where children were 19 months or older and produce more than 96 words & exclude WS forms where children were 30 months or older and produce more than 564 words) ($n_{youngest}$ = 126,$n_{middle}$ = 38 ).
4. The youngest group was filtered to not include any children from the middle group (n = 117).
5. Finally, the middle and youngest groups were filtered to only include children with multiple CDIs between 9 and 15 months apart, or single CDI administrations ($n_{youngest}$ = 108,$n_{middle}$ = 38). The table below shows the distribution of multiple CDIs across groups. 



| group | n_entry | count | mean CDI latency |
|-------|---------|-------|------------------|
|  yng  |    1    |   102 | 11.17            |       
|  yng  |    2    |   6   | 11.17            |
|  mdl  |    1    |   28  | 12.4             |
|  mdl  |    2    |   10  | 12.4             |
|  old  |    1    |   33  | 12.5             | 
|  old  |    2    |   18  | 12.5             |





```{r}
ggplot(all_subs,aes(x = nProduced,color = group))+
  geom_density()+
  theme_bw()

ggplot(all_subs, aes(x = interview_age,y=nProduced))+
  geom_point()+
  facet_grid(group~form,axes = "all_y")+
  theme_bw()
  
ggplot(all_subs, aes(x = interview_age, y = nProduced, color = form, group = subjectkey))+
  geom_point() +
  geom_line()+
  facet_grid(~group,scales = "free")+
  theme_bw()


```




