---
title: "Naigles Integrated"
author: "Stanley West"
date: "2025-11-11"
output: pdf_document
---

```{r,echo=FALSE}
library(tidyverse)
library(ggplot2)
library(purrr)
library(readxl)

letty_data <- read_xlsx("data/CDI_asd.xlsx",skip = 1) %>%
  pivot_longer(cols = ends_with(c("age","produced")),
               values_to = "value",
               names_to = c("visit","var"),
               names_sep = "_") %>%
  pivot_wider(names_from = "var",
              values_from = "value") %>%
  rename(nProduced = produced,
         subjectkey = ID,
         interview_age = age) %>%
  group_by(subjectkey) %>%
  mutate(visit_latency = interview_age -lag(interview_age,n = 2),
         source = "niagles",
         form = NA,
         group = ifelse(interview_age >= 24 & interview_age < 48, "yng", "mdl")) %>%
  select(-visit)

```


## Plot language trajectories
There are 4 kiddos with no growth throughout the three visits. There are no recordings that meet our previous 9 - 15 month latency time window. The furthest latencies (from Visit 1 to Visit 3 range from 7 - 9 months apart). None of the Niagles kids meet age criteria for being in the "old" group.

```{r}

ggplot(letty_data,aes(x = age, y = nProduced, color = subjectkey))+
  geom_point()+
  geom_line(aes(group = subjectkey))+
  theme(legend.position = "none")

## taking out low producers over all visits - 4 had recorded no language throughout visits.

ggplot(letty_data %>%
         group_by(subjectkey) %>%
         filter(!all(nProduced == 0)),aes(x = age, y = nProduced, color = subjectkey))+
  geom_point()+
  geom_line(aes(group = subjectkey))+
  theme(legend.position = "none")

summary(letty_data)

hist(letty_data$visit_latency)

```

## If we want to use this data in the context of a non-longitudinal analysis.


```{r,echo=FALSE}
ASD_cdi <- read.csv("data/ASD_CDI_subs.csv") %>%
  mutate(source = "CDI") 

merged_data <- rbind(ASD_cdi, letty_data %>%
                       select(-visit_latency) %>%
                       filter(!is.na(interview_age)))%>%
  mutate(group = factor(group, levels = c("yng", "mdl", "old"), labels =c("yng", "mdl", "old") ))

```


## Looking at unmatched distributions

```{r}
ggplot(merged_data, aes(x = nProduced))+
  geom_histogram(stat = "density")+
  facet_grid(group~form)+
  theme_bw()


ggplot(merged_data, aes(x = interview_age,y=nProduced))+
  geom_point()+
  facet_grid(group~form,axes = "all_y")+
  theme_bw()

ggplot(merged_data, aes(x =nProduced,color = group))+
  geom_density()+
  theme_bw()

ggplot(merged_data, aes(x =nProduced,color = group))+
  geom_density()+
  theme_bw()+
  facet_grid(~source)


merged_data |>
  group_by(group) |>
  summarize(n = n(),
            mean_production = mean(nProduced),
            mean_age = mean(interview_age)) |>
  kable(caption = "Descriptive Stats for Unmatched Sample")
```





