---
title: "Naigles Integrated"
author: "Stanley West"
date: "2025-11-11"
output: pdf_document
---

```{r,echo=FALSE}
library(tidyverse)
library(ggplot2)
library(purrr)
library(readxl)
library(TriMatch)
library(kableExtra)
library(MatchIt)

letty_data <- read_xlsx("data/CDI_asd.xlsx",skip = 1) %>%
  pivot_longer(cols = ends_with(c("age","produced")),
               values_to = "value",
               names_to = c("visit","var"),
               names_sep = "_") %>%
  pivot_wider(names_from = "var",
              values_from = "value") %>%
  rename(nProduced = produced,
         subjectkey = ID,
         interview_age = age) %>%
  group_by(subjectkey) %>%
  mutate(visit_latency = interview_age -lag(interview_age,n = 2),
         source = "niagles",
         form = NA,
         group = ifelse(interview_age >= 24 & interview_age < 48, "yng", "mdl")) %>%
  select(-visit)

```


## Plot language trajectories
There are 4 kiddos with no growth throughout the three visits. There are no recordings that meet our previous 9 - 15 month latency time window. The furthest latencies (from Visit 1 to Visit 3 range from 7 - 9 months apart). None of the Niagles kids meet age criteria for being in the "old" group.

```{r}

ggplot(letty_data,aes(x = interview_age, y = nProduced, color = subjectkey))+
  geom_point()+
  geom_line(aes(group = subjectkey))+
  theme(legend.position = "none")

## taking out low producers over all visits - 4 had recorded no language throughout visits.

ggplot(letty_data %>%
         group_by(subjectkey) %>%
         filter(!all(nProduced == 0)),aes(x = interview_age, y = nProduced, color = subjectkey))+
  geom_point()+
  geom_line(aes(group = subjectkey))+
  theme(legend.position = "none")

summary(letty_data)

hist(letty_data$visit_latency)

```

## If we want to use this data in the context of a non-longitudinal analysis.


```{r,echo=FALSE}
ASD_cdi <- read.csv("data/ASD_CDI_subs.csv") %>%
  mutate(source = "CDI") 

merged_data <- rbind(ASD_cdi, letty_data %>%
                       select(-visit_latency) %>%
                       filter(!is.na(interview_age)))%>%
  mutate(group = factor(group, levels = c("yng", "mdl", "old"), labels =c("yng", "mdl", "old") ))

```


## Looking at unmatched distributions

```{r}
ggplot(merged_data, aes(x = nProduced))+
  geom_histogram(stat = "density")+
  facet_grid(group~form)+
  theme_bw()


ggplot(merged_data, aes(x = interview_age,y=nProduced))+
  geom_point()+
  facet_grid(group~form,axes = "all_y")+
  theme_bw()

ggplot(merged_data, aes(x =nProduced,color = group))+
  geom_density()+
  theme_bw()

ggplot(merged_data, aes(x =nProduced,color = group))+
  geom_density()+
  theme_bw()+
  facet_grid(~source)


merged_data |>
  group_by(group) |>
  summarize(n = n(),
            mean_production = mean(nProduced),
            mean_age = mean(interview_age)) |>
  kable(caption = "Descriptive Stats for Unmatched Sample")
```


## Matching with new data

# TriMatch

```{r eval=FALSE, include=FALSE}
## Setting matching formula
formula <- ~ nProduced
merged_data$treats <- ifelse(merged_data$group == "mdl", "Control",
                          ifelse(merged_data$group == "old", "Treat1","Treat2"))
prop_scores <- trips(merged_data, merged_data$treats, formula) 

plot(prop_scores) + ggtitle("Triangle plot",subtitle = "Shows propensity scores from the three models. Each subject has two scores,\n connected by  a line.")

```
## Distribution and Descriptives with a 0.15 SD caliper cutoff. 

```{r eval=FALSE, include=FALSE}

prop_scores <- prop_scores|>
  mutate(subjectkey = merged_data$subjectkey)

all_data.matched <- trimatch(prop_scores, method = NULL, caliper = 0.15,nmatch = Inf) |>
  pivot_longer(cols = c("Treat1", "Control","Treat2"),
               names_to = "treat",
               values_to = "id",
               values_transform = as.integer) |>
  left_join(select(prop_scores,subjectkey,id), by = "id")

trimatched_subs <- merged_data |>
  filter(subjectkey %in% all_data.matched$subjectkey)


ggplot(trimatched_subs, aes(x = nProduced, color = group))+
  geom_density()+
  theme_bw()

trimatched_subs |>
  group_by(group) |>
  summarize(n = n(),
            mean_age = mean(interview_age),
            sd_age = sd(interview_age),
            mean_produced = mean(nProduced),
            sd_produced = sd(nProduced)) |>
  as.data.frame()|>
  kable(col.names = c("group","n", "M(age)", "SD(age)", "M(produced)", "SD(produced)"))

```


## Series of calipers
Setting lower calipers (in standardized units) makes the matching more strict. More liberal calipers keeps larger distance propensity triplets. As you approach 0.5, you retain more and more data until you are left with the full unmatched sample at 0.5. Differences in distributions across caliper cutoffs seem to dissappear at around 0.25.


```{r eval=FALSE, include=FALSE}
series <- map_dfr(.x = seq(0.05,0.5,0.05),function(x){
  d_matched <- trimatch(prop_scores, method = NULL, caliper = x,nmatch = Inf) |>
    pivot_longer(cols = c("Treat1", "Control","Treat2"),
               names_to = "treat",
               values_to = "id",
               values_transform = as.integer) |>
  left_join(select(prop_scores,subjectkey,id), by = "id")
  
  trimatched_subs <- merged_data |>
    filter(subjectkey %in% d_matched$subjectkey) |>
    mutate(caliper = x)
  return(trimatched_subs)
})

ggplot(series , aes(x = nProduced, color = group))+
  geom_density()+
  theme_bw()+
  facet_wrap(~caliper) +
  labs(title = "Matching nProduced distributions w/ different caliper cutoffs")



series |>
  group_by(caliper,group) |>
  summarize(n = n(),
            mean_age = mean(interview_age),
            sd_age = sd(interview_age),
            mean_produced = mean(nProduced),
            sd_produced = sd(nProduced)) |>
  as.data.frame()|>
  kable(col.names = c("caliper","group","n", "M(age)", "SD(age)", "M(produced)", "SD(produced)"))

```


## Hand matching
Step-wise distributions in vocabulary, as expected. Young have high proportion of low vocabularies and low distribution of high. Middle have second highest proportion of low vocabs and a bump at intermediate vocab, and old have lowest proportion of small vocab and largest proportion of high vocabs.

```{r eval=FALSE, include=FALSE}

map(seq.int(0,10), function(x){
  yng_z <- merged_data %>%
    filter(group == "yng",nProduced <= 50) %>%
    sample(.,x,replace = FALSE)
  d <- merged_data[yng_z]
  return(d)
})



```

