---
title: "Naigles Integrated"
author: "Stanley West"
date: "2025-11-11"
output: pdf_document
---

```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(tidyverse)
library(ggplot2)
library(purrr)
library(readxl)
library(TriMatch)
library(kableExtra)
library(MatchIt)
library(emmeans)
library(rstatix)
letty_data <- read_xlsx("data/CDI_asd.xlsx",skip = 1)[-c(33:44),] %>%
  select(-V2_W_G_says) %>%
  rename(V1_age = V1_Age,
         V1_produced = V1_W_G_says,
         V2_age = V2_age,
         V2_produced = V2CDITotalwords_W_P,
         V3_age = V3Age,
         V3_produced = V3CDITotalwords_WP ) %>%
  pivot_longer(cols = ends_with(c("age","produced")),
               values_to = "value",
               names_to = c("visit","var"),
               names_sep = "_") %>%
  pivot_wider(names_from = "var",
              values_from = "value") %>%
  rename(nProduced = produced,
         subjectkey = ID,
         interview_age = age) %>%
  group_by(subjectkey) %>%
  mutate(visit_latency = interview_age -lag(interview_age,n = 2),
         source = "naigles",
         form = NA,
         group = ifelse(interview_age >= 24 & interview_age < 48, "yng", "mdl")) %>%
  select(-visit)

```


## Plot language trajectories
There are 4 kiddos with no growth throughout the three visits. There are no recordings that meet our previous 9 - 15 month latency time window. The furthest latencies (from Visit 1 to Visit 3 range from 7 - 9 months apart). None of the Niagles kids meet age criteria for being in the "old" group.

```{r}

ggplot(letty_data,aes(x = interview_age, y = nProduced, color = subjectkey))+
  geom_point()+
  geom_line(aes(group = subjectkey))+
  theme(legend.position = "none")

## taking out low producers over all visits - 4 had recorded no language throughout visits.

ggplot(letty_data %>%
         group_by(subjectkey) %>%
         filter(!all(nProduced == 0)),aes(x = interview_age, y = nProduced, color = subjectkey))+
  geom_point()+
  geom_line(aes(group = subjectkey))+
  theme(legend.position = "none")

summary(letty_data)

hist(letty_data$visit_latency)


```

## If we want to use this data in the context of a non-longitudinal analysis.


```{r,echo=FALSE}
ASD_cdi <- read.csv("data/ASD_CDI_subs.csv") %>%
  mutate(source = "CDI") 

merged_data <- rbind(ASD_cdi, letty_data %>%
                       select(-visit_latency) %>%
                       filter(!is.na(interview_age)))%>%
  mutate(group = factor(group, levels = c("yng", "mdl", "old"), labels =c("yng", "mdl", "old") ))

```


## Looking at unmatched distributions
New data only fits into the young and middle groups. 

```{r}


ggplot(merged_data, aes(x =nProduced,color = group))+
  geom_density()+
  theme_bw()+
  ggtitle("Combined Data Distributions")

ggplot(merged_data, aes(x =nProduced,color = group))+
  geom_density()+
  theme_bw()+
  facet_grid(~source)+
  ggtitle("Separate Data Spreads by Source")


merged_data |>
  group_by(group) |>
  summarize(n = n(),
            mean_production = mean(nProduced),
            mean_age = mean(interview_age)) |>
  kable(caption = "Descriptive Stats for Unmatched Sample")
```


## Matching single time points. Two groups (non-minimally speaking and minimally speaking) w/ added Naigles data

```{r}
twog <- merged_data %>%
  mutate(min_speak = ifelse(group == "old", TRUE,FALSE))


match_m_twog <- matchit(min_speak ~ nProduced,
                        data = twog,
                        method = "optimal",
                        distance = "glm")
summary(match_m_twog)

plot(match_m_twog,type = "density")

match_twog <- merged_data %>%
  right_join(match.data(match_m_twog), by = c("subjectkey","interview_age","nProduced","group","source","form"))

t.test(match_twog$nProduced[match_twog$min_speak == TRUE],match_twog$nProduced[match_twog$min_speak == FALSE] )


ggplot(match_twog, aes(x = nProduced, color = min_speak))+
  geom_density()


ggplot(match_twog, aes(x = interview_age,y = nProduced, color = source,shape=min_speak))+
  geom_point()


```

## Matching single time points. Two groups (non-minimally speaking and minimally speaking) w/o added Naigles data
We get a slightly better match not using the Naigles data in a two-group solution. $p = 0.968$ vs. $p = 0.957$
```{r}
twog <- merged_data %>%
  mutate(min_speak = ifelse(group == "old", TRUE,FALSE)) %>%
  filter(!source == "naigles")


match_m_twog <- matchit(min_speak ~ nProduced,
                        data = twog,
                        method = "optimal",
                        distance = "glm")
summary(match_m_twog)

plot(match_m_twog,type = "density")

match_twog <- merged_data %>%
  right_join(match.data(match_m_twog), by = c("subjectkey","interview_age","nProduced","group","source","form"))

t.test(match_twog$nProduced[match_twog$min_speak == TRUE],match_twog$nProduced[match_twog$min_speak == FALSE] )


ggplot(match_twog, aes(x = nProduced, color = min_speak))+
  geom_density()


ggplot(match_twog, aes(x = interview_age,y = nProduced, shape = min_speak))+
  geom_point()


```


## Three time points -
Hand matching and batch-matching. Took the 30 smallest vocabularies out of the youngest group and the top 5 vocabularies from the middle group. The rest of the matching was done by iterativly removing specific vocabularies from the data to better match across the whole range of vocabularies. One-way ANOVA reveals no significant group differences in number of words produced $p = 0.773$

```{r}
yng_bump <- merged_data %>%
  filter(nProduced >= 3) %>%
  filter(group == "yng") %>%
  slice_min(nProduced, n =30)
yng_repeats <- merged_data %>%
  filter(nProduced >= 3) %>%
  filter(source == "naigles") %>%
  group_by(subjectkey) %>%
  slice_max(nProduced,n=1)
mdl_bump <-merged_data %>%
  filter(nProduced >= 3) %>%
  filter(group == "mdl") %>%
  slice_max(nProduced, n =5) 

new_match <- merged_data %>%
  filter(nProduced >=3) %>%
  filter(!subjectkey %in% yng_bump$subjectkey,
         !subjectkey %in% mdl_bump$subjectkey,
         !subjectkey %in% yng_repeats$subjectkey,
         !subjectkey == "NDARPR371BH1",
         !subjectkey == "NDAR_INVJV336XCL",
         !subjectkey == "NDARFH331TLN",
         !subjectkey == "NDAR_INVAC323FPZ",
         !subjectkey == "NDARAB989DG5",
         !subjectkey =="NDAR_INVCC993TBL",
         !subjectkey == "NDARVL893WB0",
         !subjectkey == "NDARHX114VE2",
         !subjectkey == "NDARVV356TKH",
         !subjectkey == "NDARVG708MPB",
         !subjectkey == "ASD_3_o",
         !subjectkey == "NDARAW094GEC",
         !subjectkey == "ASD_3_l",
         !subjectkey == "ASD_2_e",
         !subjectkey == "NDAR_INVEC869JNQ",
         !subjectkey == "NDARAZ926NXF",
         !subjectkey == "ASD_2_i",
         !subjectkey == "NDARVZ466UF3") %>%
  group_by(subjectkey) %>%
  mutate(n_entry = length(subjectkey)) %>%
  rbind(yng_repeats)



ggplot(new_match, aes(x =nProduced,color = group))+
  geom_density()+
  theme_bw()
summary(new_match)
summary(aov(nProduced ~ group, data = new_match))

new_match %>%
  group_by(group) %>%
  summarize(n())

new_match %>%
  group_by(group,subjectkey) %>%
  summarize(n = n()) %>%
  arrange(desc(n))

emmeans_test(nProduced ~ group, 
             data = new_match %>% ungroup(),
             p.adjust.method = "bonferroni",
             detailed = FALSE)


```






