---
title: "New GUIDS Breakdown"
author: "Stan West"
date: "2025-04-16"
output: slidy_presentation
theme: "lumen"
---

```{r setup, include=FALSE}
library(ggplot2)
library(tidyverse)
library(readxl)
library(knitr)
load("data/ASD_all_GM.Rdata")

sentences <- read.csv("data/mci_sentences02.csv",na.strings = "")[-1,]
gestures <- read.csv("data/mci_words_gestures01.csv",na.strings = "")[-1,]
checked <- read_xlsx("data/GUIDS_to_check_April2023.xlsx")
checked2 <- read_xlsx("data/GUIDs_N200_June_2022_SRCLD.xlsx")
growthmod <- read.csv("data/ASD_GUIDS.csv")[,-1]
new_ids <-read_xlsx("data/GUIDS_new_March_2025_aut_dx_confirmed.xlsx")

word_dict_S <- read.csv("data/word_dict_S_April2023.csv")
names(word_dict_S) <- c('num_item_id', 'word_code', 'word1', 'word2', 'WordBank1', 'WordBank2', 'category', 'class', 'classForm', 'classEH', 'classRS','classEHse','classVerbs', 'CDI_Metadata_compatible')
word_dict_WG <- read.csv("data/word_dict_WG.csv")
names(word_dict_WG) <- c('num_item_id','word_code', 'word1', 'word2', 'WordBank1', 'WordBank2', 'category', 'class', 'classForm', 'classEH', 'classRS', 'classEHse','classVerbs', 'CDI_Metadata_compatible')





## 2. Selecting vocabulary checkists and other demographic info. ########################################
select_columns <- c('subjectkey', 'interview_age', 'sex','interview_date','mcs_vc_total','mci_sentences02_id', as.character(word_dict_S$word_code))
mci_sentences02 <- sentences[select_columns] %>%
  filter(!duplicated(.)) %>%
  mutate(form = "WS",
         interview_age = as.integer(interview_age),
         interview_date = as.Date(interview_date, "%m/%d/%Y"),.after="sex")

select_columns <- c('subjectkey', 'interview_age', 'sex','interview_date','mcg_vc_totpr','mci_words_gestures01_id', as.character(word_dict_WG$word_code))
mci_words_gestures <- gestures[select_columns] %>%
  filter(!duplicated(.))%>%
  mutate(form = "WG",
         interview_age = as.integer(interview_age),
         interview_date = as.Date(interview_date, "%m/%d/%Y"),.after = "sex")

source("R/Identify_empty_checklist.R")
source("R/adds_nProduced_column.R")

```

# Check Pseudo IDs

-   No overlap in multiple GUIDs with PseudoIDs

```{r}

new_ids %>% 
  select(Pseudoguid, subjectkey.GUID) %>%
  filter(!is.na(Pseudoguid)) 

```

# New Kids w/vocab 1 - 680 (VSOA Analysis)

- 250 new CDIs coming from 122 unique GUIDs

```{r, echo=FALSE}
gestures_no_na <- mci_words_gestures %>%
  mutate(empty = Identify_empty_checklist(.[,-c(1:6)], 0.9),
         mcg_vc_totpr = as.numeric(mcg_vc_totpr),
         nproduced = adds_nProduced_column(d=.,start_at = 7, 2),.after = mcg_vc_totpr) %>%
  filter(empty == FALSE,
         nproduced > 0,
         subjectkey %in% new_ids$subjectkey.GUID)



sentences_no_na <- mci_sentences02 %>%
  mutate(empty = Identify_empty_checklist(.[,-c(1:6)], 0.9),
         mcs_vc_total = as.numeric(mcs_vc_total),
         nproduced = adds_nProduced_column(.,7,1,3),.after=mcs_vc_total) %>%
  filter(empty == FALSE,
         nproduced > 0,
         subjectkey %in% new_ids$subjectkey.GUID) 


df_plot <- rbind(
  gestures_no_na %>% select(subjectkey,
                            nproduced,
                            form,
                            interview_age),
  sentences_no_na %>% select(subjectkey,
                            nproduced,
                            form,
                            interview_age)
) %>%
  unique() %>%
  arrange(subjectkey,interview_age)%>%
  group_by(subjectkey,interview_age) %>%
  slice_max(nproduced)



ggplot(df_plot, aes(y = nproduced,x = interview_age, color = form))+
  geom_point()+
  facet_grid(~form,scales = "free")+
  lims(y = c(1,700))+
  labs(x = "Interview Age",
       y = "# of words produced",
       title = "Words Produced by Age Scatterplot",
       caption = "Points are individual CDIs")


```

# Kids between 3 and 650 words w/ longitudinal (~12 months apart) data
- 557 CDIs, 337 unique GUIDs
- 43 kids were identified as having vocabulary data ~12 months apart
- Descriptives of the maximum differences within participants are displayed in the table

\n

```{r, echo=FALSE,message=FALSE,warning=FALSE}
gestures_no_na_650 <- mci_words_gestures %>%
  mutate(empty = Identify_empty_checklist(.[,-c(1:6)], 0.9),
         mcg_vc_totpr = as.numeric(mcg_vc_totpr),
         nproduced = adds_nProduced_column(d=.,start_at = 7, 2),.after = mcg_vc_totpr) %>%
  filter(empty == FALSE,
         nproduced > 3 & nproduced < 650,
         subjectkey %in% new_ids$subjectkey.GUID)



sentences_no_na_650 <- mci_sentences02 %>%
  mutate(empty = Identify_empty_checklist(.[,-c(1:6)], 0.9),
         mcs_vc_total = as.numeric(mcs_vc_total),
         nproduced = adds_nProduced_column(.,7,1,3),.after=mcs_vc_total) %>%
  filter(empty == FALSE,
         nproduced > 3 & nproduced < 650,
         subjectkey %in% new_ids$subjectkey.GUID) 

df_plot_650 <- rbind(
  gestures_no_na_650 %>% select(subjectkey,
                            nproduced,
                            form,
                            interview_age),
  sentences_no_na_650 %>% select(subjectkey,
                            nproduced,
                            form,
                            interview_age)
) %>%
  unique() %>%
  arrange(subjectkey,interview_age)%>%
  group_by(subjectkey,interview_age) %>%
  slice_max(nproduced) 


ASD_wg_long <- gestures_no_na_650 %>%
  pivot_longer(.,
               cols = matches("^mcg_vc?[0-9]"),
               values_to = "response_code",
               names_to = "word_code") %>%
  mutate(response_code = as.numeric(response_code),
         interview_age = as.numeric(interview_age),
        mci_words_gestures01_id = as.numeric(mci_words_gestures01_id)) %>%
  mutate_at(vars("response_code"), ~replace_na(.,0)) %>%
  mutate(Produces = ifelse(response_code == 2, TRUE, FALSE)) %>%
  group_by(subjectkey, interview_age, interview_date, mci_words_gestures01_id) %>%
  mutate(nProduced = sum(Produces)) %>%
  filter(nProduced >= 20 & nProduced <= 600) %>%
  left_join(.,word_dict_WG, by = "word_code") %>%
  ungroup() %>%
  select(-mci_words_gestures01_id,  -mcg_vc_totpr)

ASD_ws_long <- sentences_no_na_650 %>%
  pivot_longer(.,
               cols = matches("^mcs_vc?[0-9]"),
               values_to = "response_code",
               names_to = "word_code") %>%
  mutate(response_code = as.numeric(response_code),
         interview_age = as.numeric(interview_age),
         interview_date = as.Date(interview_date, form = "%m/%d/%Y")) %>%
  mutate_at(vars("response_code"), ~replace_na(.,0)) %>%
  mutate(Produces = ifelse(response_code == 1 | response_code == 3, TRUE, FALSE)) %>%
  group_by(subjectkey, interview_age, interview_date, mci_sentences02_id) %>%
  mutate(nProduced = sum(Produces)) %>%
  filter(nProduced >= 20 & nProduced <= 600) %>%
  left_join(.,word_dict_S, by = "word_code") %>%
  ungroup() %>%
  select(-mci_sentences02_id, -mcs_vc_total)


ASD_all_long <- rbind(ASD_wg_long, ASD_ws_long) %>%
  group_by(subjectkey) %>%
  mutate(best = nProduced == max(nProduced),
         group = "ASD") %>%
  filter(best == TRUE)%>%
  ungroup() %>%
  group_by(subjectkey) %>%
  mutate(newest_rec = interview_date == max(interview_date)) %>%
  group_by(subjectkey, interview_age) %>%
  filter(newest_rec == TRUE) %>%
  ungroup() %>%
  unique() %>%
  select(-nproduced,-best,-group,-newest_rec)

ASD_total <- rbind(ASD_all_long,ASD_all)


df_plot_650_diffs_total <- ASD_total %>% select(subjectkey,
                            nProduced,
                            form,
                            interview_age) %>%
  unique() %>%
  arrange(subjectkey,interview_age)%>%
  group_by(subjectkey,interview_age) %>%
  slice_max(nProduced) 

diffs <- df_plot_650_diffs_total %>%
  select(subjectkey,interview_age) %>%
  group_by(subjectkey) %>%
  mutate(entry = seq.int(1,n(),1)) %>%
  filter(!max(entry) < 2) %>%
  mutate(diff = interview_age - lag(interview_age),
         max_diff = max(diff,na.rm = T))

tbl_diffs <- diffs %>%
  ungroup() %>%
  select(subjectkey,max_diff) %>%
  unique() %>%
  filter(max_diff >= 9 & max_diff <= 15) %>%
  summarize(n = length(max_diff),
            range = range(max_diff),
            mean = mean(max_diff),
            sd = sd(max_diff))

ggplot(diffs %>% filter(!is.na(diff)), aes(x = diff))+
  geom_histogram(fill = "black")+
  labs(x = "Distribution of Time Between Interviews",
       title = "Interview Age Distances Histogram (no distance constraint)")

kable(tbl_diffs,caption = "Maximum Interview Latencies Descriptives",format = "simple")

ggplot(diffs %>% 
         filter(max_diff >= 9 & max_diff <= 15) %>%
         left_join(.,
                   ASD_total %>%
                     select(subjectkey,
                            interview_age,
                            nProduced), by = c("subjectkey", "interview_age")), 
       aes(y = nProduced, x = interview_age))+
  geom_point(aes(color = subjectkey))+
  geom_line(aes(color = subjectkey))+
  theme_bw()+
  theme(legend.position = "none")+
  labs(x = "Interview Age",
       y = "# of words produced",
       title ="Longitudinal Relationship Between Age and Word Production",
       caption = "Differences of between 9 and 15 months considered\n Colors indicate different kids")


x <- diffs %>%
  filter(diff >= 2 | diff <= 4) %>%
  group_by(subjectkey) %>%
  summarize(n = length(diff)) %>%
  arrange(desc(n)) %>%
  filter(n > 1)


## Mean nProduced per kid.
diffs %>%
  filter(diff >= 2 | diff <= 4) %>%
  group_by(subjectkey) %>%
  left_join(., ASD_total %>% select(subjectkey, nProduced)) %>%
  summarize(m = mean(nProduced))

x %>%
  group_by(n) %>%
  summarize(n())

length(unique(x$subjectkey))

ggplot(x, aes(x = n))+
  geom_histogram()

```

```{r,echo=FALSE}
minimal <- df_plot %>%
  filter(interview_age >= 60) %>%
  left_join(.,new_ids %>% select(subjectkey.GUID,ADOS.Mod.1), by = c("subjectkey" = "subjectkey.GUID"))

write.csv(minimal,"minimally_speaking.csv")
```



```{r}
sentences_VSOA <- sentences_no_na %>%
  filter(subjectkey %in% df_plot$subjectkey & 
           interview_age %in% df_plot$interview_age )

gestures_VSOA <- gestures_no_na %>%
  filter(subjectkey %in% df_plot$subjectkey & 
           interview_age %in% df_plot$interview_age )
```




# Filter 3 to 600 CDIs and kids

```{r}
ASD_total %>%
  filter(nProduced > 2 & nProduced <= 600) %>%
  select(interview_age, subjectkey) %>%
  unique() %>%
  nrow()

ASD_total %>%
  filter(nProduced > 2 & nProduced <= 600) %>%
  select(subjectkey, nProduced) %>%
  summarize(m = mean(nProduced))


  

ASD_3_600 <- ASD_total %>% select(subjectkey,
                            nProduced,
                            form,
                            interview_age) %>%
  unique() %>%
  filter(nProduced > 2 & nProduced <= 600) %>%
  arrange(subjectkey,interview_age)%>%
  group_by(subjectkey,interview_age) %>%
  slice_max(nProduced) %>%
  select(subjectkey,interview_age) %>%
  group_by(subjectkey) %>%
  mutate(entry = seq.int(1,n(),1),
         max_entry = max(entry)) %>%
  filter(max_entry >= 2) %>%
  mutate(diff = interview_age - lag(interview_age),
         max_diff = max(diff,na.rm = T)) %>%
  filter(diff >= 9 & diff <= 15)


ASD_3_600 %>%
  group_by(subjectkey) %>%
  summarize(num_entries = length(entry)) %>%
  group_by(num_entries) %>%
  summarize(length(num_entries))
  



ggplot(ASD_3_600 %>% 
         left_join(.,
                   ASD_total %>%
                     select(subjectkey,
                            interview_age,
                            nProduced), by = c("subjectkey", "interview_age")), 
       aes(y = nProduced, x = interview_age))+
  geom_point(aes(color = subjectkey))+
  geom_line(aes(color = subjectkey))+
  theme_bw()+
  theme(legend.position = "none")+
  labs(x = "Interview Age",
       y = "# of words produced",
       title ="Longitudinal Relationship Between Age and Word Production",
       caption = "Differences of between 9 and 15 months considered\n Colors indicate different kids")


minmial_all <- ASD_total %>%
  filter(interview_age >= 60) %>%
  select(subjectkey, interview_age, form, nProduced)
  unique()

write.csv(minmial_all, "data/minimal_ASD_all.csv")


save(ASD_total, file = "data/ASD_longitudinal_all.Rdata")

```



